Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-26 submission/manuscript.Rmd
discussion edit

In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the adult _Kdm5c_-KO hippocampus and amygdala. We defined tissue-enriched genes based on the criteria used in a previously published dataset of 17 C57/Bl6J mouse tissues[@liComprehensiveMouseTranscriptomic2017], which defined genes as tissue-enriched if they had more than 4-fold higher expression when compared to any other tissue. In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. It is currently unclear what the impact of these tissue-enriched genes are upon _Kdm5c_-KO brain function and if they ultimately contribute to MRXSCJ-related phenotypes. While the majority of tissue-enriched DEGs were testis and ovary genes with no known brain functions, select liver and muscle-biased DEGs do have known roles in the wild-type brain. For example, _Apolipoprotein C-I (Apoc1)_ is a lipid transport gene highly enriched within the liver, but is also lowly expressed in the wild-type brain. Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010]. Our results suggest KDM5C fine-tunes the expression of tissue-enriched genes like _Apoc1_ to match the level requried for proper brain function. However, futher studies are required to detrmine if these genes contribute to _Kdm5c_-KO neuronal impairments and MRXSCJ pathology.

>>>>

In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the adult _Kdm5c_-KO hippocampus and amygdala. In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. Given the majority of tissue-enriched DEGs are testis and ovary genes with no known brain functions, it is currently unknown if they impair _Kdm5c_-KO brain function and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression. However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus. Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010]. Thus, failure to fine-tune the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_ could be one route by which loss of brain tissue identity contributes to _Kdm5c_-KO impairments.
